### 2015-07-23 Harris_2015/manuscript-materials/draft.Rmd
TeX

**Abstract**

Estimating species interactions from observational data is one of the most
controversial tasks in community ecology. One difficulty is that a single
pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, two competing species would
ordinarily correlate negatively in space, but this effect can be reversed in
the presence of a third species that is capable of outcompeting both of them
when it is present. Here, I apply models from statistical physics, called
Markov networks or Markov random fields, that can predict the direct and
indirect consequences of any possible species interaction matrix. Interactions
in these models can be estimated from observational data via maximum
likelihood. Using simulated landscapes with known pairwise interaction
strengths, I evaluated Markov networks and several existing approaches. The
Markov networks consistently outperformed other methods, correctly isolating
direct interactions between species pairs even when indirect interactions or
abiotic environmental effects largely overpowered them. Two computationally
efficient approximations, based on linear and generalized linear models, also performed
well. Indirect effects reliably caused a common null modeling approach and a Joint
Species Distribution Modeling approach to produce incorrect inferences, however.

@@

\setlength{\parindent}{1cm}

\noindent  
**Introduction**

@@

\noindent  **Methods**

\noindent  
**Conditional relationships and Markov networks.**
Ecologists are often interested in learning about direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate *conditional*
("all-else-equal") relationships, rather than *marginal* ("overall") relationships.
The example with the shrubs and trees in
Figure 1 shows how the two measures can have opposite signs,
and suggests that conditional relationships are more relevant
for drawing inferences about species interactions (e.g. competition).

@@

$$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).$$

@@

\noindent  
**Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.**
In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
however, ecologists will often need to estimate these parameters from co-occurrence data
(i.e. from a matrix of ones and zeros indicating which species are present at which sites).
When the number of species is reasonably small, one can compute exact maximum
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients given a
presence-absence matrix by optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. This maximum represents the unique combination of $\alpha$ and $\beta$
coefficients that would be expected to produce exactly the observed co-occurrence
frequencies (i.e. it matches the sufficient statistics of the model distribution to the
sufficient statistics of the data). For the analyses in this paper, I used the *rosalia* package
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
define the objective function and gradient in R code and then find the optimum.

\noindent  
**Simulated landscapes.**
In order to compare different methods, I simulated two sets of landscapes using known
parameters.

@@

\noindent  
**Recovering species interactions from simulated data.**
I used Markov networks and  several other techniques for
determining the sign and strength of the associations between
pairs of species (Appendix B).

@@

The third alternative, generalized linear models, can also be thought of as a computationally
efficient approximation to the Markov network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit regularized logistic regression models for each species, using
the other species on the landscape as predictors. [[Observational presence-absence data does
not have enough degrees of freedom to estimate this many parameters, so I symmetrized the
pairwise correlations by averaging the coefficient for predicting species $i$’s status from
species $j$ and vice versa.]]

@@

The last two methods used the latent correlation matrix estimated by the BayesComm
package [@REF] to evaluate the claim from recent papers that the correlation
coefficients estimated by “joint species distribution models” provide an accurate
assessment of species’ pairwise interactions [@pollock_understanding_2014;
see also @harris_generating_2015].  I reported both the average correlation
between each pair of species’ latent variables across 1000 Monte Carlo
samples, as well as the average partial correlation.

\noindent  
**Evaluating model performance.**
For the simulated landscapes based on Figure 1, method evaluation was fairly
qualitative: any method whose test statistic for the two shrubs indicated a
negative relationship passed; other methods failed. For methods that provide
confidence intervals or other inferential statistics (Appendix), I also reported how
often they rejected the null hypothesis of no interaction.

@@

\noindent  
**Results**

\noindent  
**Three species.**
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the
covariance method falsely reported positive associations for 94% of the simulated
landscapes, and the randomization-based null model falsely reported such
associations 100% of the time. The [[two]] methods for evaluating conditional
relationships (Markov networks and partial covariances), however, successfully
controlled for the indirect pathway via the tree species and each correctly
identified the direct negative interaction between the shrubs 94% of the time.

\noindent  
**Twenty species.**
The accuracy of the four evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3). In general,
however, there was a consistent ordering…

@@

\noindent  
**Discussion**

\noindent  
The results presented above show that Markov networks can reliably recover
species' pairwise interactions from observational data, even for cases where
a common null modeling technique reliably fails. Specifically, Markov networks
were successful even when direct interactions were largely overwhelmed by indirect
effects (Figure 1). For cases where fitting a Markov network is computationally
infeasible, these results also indicate that partial covariances and generalized linear
 models (the two methods that estimated conditional relationships rather than
marginal ones) can both provide useful approximations. The partial correlations'
success on simulated data may not carry over to real data sets, however;
@loh_structure_2013 show that the linear approximations can be less reliable
in cases where the true interaction matrix contains more structure (e.g. guilds
or trophic levels). Similarly, the approximation involved in using separate
generalized linear models for each species can occasionally lead to catastrophic
overfitting with small-to-moderate sample sizes [@lee_learning_2012].
For these reasons, it will usually be best to fit a Markov network
rather than one of the alternative methods when one’s computational resources
allow it.

@@

Despite these limitations, Markov networks have enormous potential to improve
ecological understanding. For example, Markov networks provide a simple answer
to the question of how competition should affect a species' overall
prevalence, which was a major flashpoint for the null model debates in the
1980’s [@roughgarden_competition_1983; @strong_ecological_1984]. Equation
1 can be used to calculate the expected prevalence of a species in the
absence of biotic influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012].
Competition's effect on prevalence in a Markov network can then be calculated
by subtracting this value and the observed prevalence (cf Figure 2D).  Apart from
BayesComm (which had the worst performance of any method evaluated here), none of
the other methods could produce this insight because they do not make predictions
about species composition. [[*back this up with more evidence or remove it*]]

@@

\noindent  
**Acknowledgements:**
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, N. J. Gotelli, and C. F. Dormann provided useful feedback on earlier drafts
of this work.

\setlength{\parindent}{0cm}

>>>>

**Abstract:** Estimating species interactions from observational data is one of the most
controversial tasks in community ecology. One difficulty is that a single
pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, two competing species would
ordinarily correlate negatively in space, but this effect can be reversed in
the presence of a third species that is capable of outcompeting both of them
when it is present. Here, I apply models from statistical physics, called
Markov networks or Markov random fields, that can predict the direct and
indirect consequences of any possible species interaction matrix. Interactions
in these models can be estimated from observational data via maximum
likelihood. Using simulated landscapes with known pairwise interaction
strengths, I evaluated Markov networks and several existing approaches. The
Markov networks consistently outperformed other methods, correctly isolating
direct interactions between species pairs even when indirect interactions or
abiotic environmental effects largely overpowered them. Two computationally
efficient approximations, based on linear and generalized linear models, also
performed well. Indirect effects reliably caused a common null modeling approach
to produce incorrect inferences, however.

@@

\setlength{\parindent}{1cm}
\setlength{\parskip}{6pt}

### Introduction

@@

### Methods

#### Conditional relationships and Markov networks.
Ecologists are often interested in learning about direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate *conditional*
("all-else-equal") relationships, rather than *marginal* ("overall") relationships.
The example with the shrubs and trees in
Figure 1 shows how the two measures can have opposite signs,
and suggests that conditional relationships are more relevant
for drawing inferences about species interactions (e.g. competition).

@@

\centering
$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).$
\raggedright

@@

#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
however, ecologists will often need to estimate these parameters from co-occurrence data
(i.e. from a matrix of ones and zeros indicating which species are present at which sites).
When the number of species is reasonably small, one can compute exact maximum
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients given a
presence-absence matrix by optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. This maximum represents the unique combination of $\alpha$ and $\beta$
coefficients that would be expected to produce exactly the observed co-occurrence
frequencies (i.e. it matches the sufficient statistics of the model distribution to the
sufficient statistics of the data). For the analyses in this paper, I used the *rosalia* package
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
define the objective function and gradient in R code and then find the optimum.

#### Simulated landscapes.
In order to compare different methods, I simulated two sets of landscapes using
known parameters.

@@

#### Recovering species interactions from simulated data.
I used Markov networks and  several other techniques for
determining the sign and strength of the associations between
pairs of species (Appendix B).

@@

The third alternative, generalized linear models (GLMs), can also be
thought of as a computationally efficient approximation to the Markov
network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit
regularized logistic regression models for each species, using
the other species on the landscape as predictors. [[Observational
presence-absence data does not have enough degrees of freedom to estimate this
many parameters, so I symmetrized the pairwise correlations by averaging the
coefficient for predicting species $i$’s status from species $j$ and vice
versa.]]

@@

The last two methods used the latent correlation matrix estimated by the
BayesComm package [@REF] to evaluate the claim from recent papers that the
correlation coefficients estimated by “joint species distribution models”
provide an accurate assessment of species’ pairwise interactions
[@pollock_understanding_2014; see also @harris_generating_2015].  I reported
both the average correlation between each pair of species’ latent variables
across 1000 Monte Carlo samples, as well as the average *partial* correlation.

#### Evaluating model performance.
For the simulated landscapes based on Figure 1, method evaluation was fairly
qualitative: any method whose test statistic for the two shrubs indicated a
negative relationship passed; other methods failed. For methods that provide
confidence intervals or other inferential statistics (Appendix), I also reported
how often they rejected the null hypothesis of no interaction.

@@

### Results

#### Three species.
As shown in Figure 1, the marginal relationship between the two shrub species
was positive---despite their competition for space at a mechanistic level---due
to indirect effects of the dominant tree species. As a result, the
covariance method falsely reported positive associations for 94% of the simulated
landscapes, and the randomization-based null model falsely reported such
associations 100% of the time. The methods for evaluating conditional
relationships (Markov networks, GLMs, and both types of partial covariance
estimates), however, successfully controlled for the indirect pathway via
the tree species and each correctly identified the direct negative interaction
between the shrubs 94% of the time.

#### Twenty species.
The accuracy of the four evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3). In general,
however, there was a consistent ordering…

@@

### Discussion

The results presented above show that Markov networks can reliably recover
species' pairwise interactions from observational data, even for cases where
a common null modeling technique reliably fails. Specifically, Markov networks
were successful even when direct interactions were largely overwhelmed by indirect
effects (Figure 1). For cases where fitting a Markov network is computationally
infeasible, these results also indicate that partial covariances and generalized linear
 models (the two methods that estimated conditional relationships rather than
marginal ones) can both provide useful approximations. The partial correlations'
success on simulated data may not carry over to real data sets, however;
@loh_structure_2013 show that the linear approximations can be less reliable
in cases where the true interaction matrix contains more structure (e.g. guilds
or trophic levels). Similarly, the approximation involved in using separate
generalized linear models for each species can occasionally lead to catastrophic
overfitting with small-to-moderate sample sizes [@lee_learning_2012].
For these reasons, it will usually be best to fit a Markov network
rather than one of the alternative methods when one’s computational resources
allow it.

@@

Despite these limitations, Markov networks have enormous potential to improve
ecological understanding. For example, Markov networks provide a simple answer
to the question of how competition should affect a species' overall
prevalence, which was a major flashpoint for the null model debates in the
1980’s [@roughgarden_competition_1983; @strong_ecological_1984]. Equation
1 can be used to calculate the expected prevalence of a species in the
absence of biotic influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012].
Competition's effect on prevalence in a Markov network can then be calculated
by subtracting this value and the observed prevalence (cf Figure 2D).

@@

#### Acknowledgements:
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, N. J. Gotelli, and C. F. Dormann provided useful feedback on earlier drafts
of this work.

\setlength{\parindent}{0cm}
\setlength{\parskip}{4pt}
